---
title: "Energy poverty in the Netherlands"
author: "Matthijs, David, Wessel, Quint, Zakaria"
date: "2026-06-08"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
require(tidyverse)
require(rmarkdown)
require(yaml)
library(tidyr)
library(writexl)
library(readxl)
```

# Part 1: Identify a Social Problem

## 1.1 Describe the Social Problem

In recent years, energy prices in the Netherlands have increased significantly, leading to a substantial rise in fixed costs of Dutch households. This increase in fixed expenses reduces the financial resources available for those household, limiting their capacity to meet other essential needs. To effectively address this growing issue among those households, it is crucial to quantify and understand the extent to which rising energy prices drive poverty rates among the Dutch households.

# Part 2: Data Sourcing

## 2.1 Load in the data

```{r}
raw_energie <- read_csv2("Energietarieven.csv")
raw_inkomen <- read_csv2("Inkomen huishoudens samengevoegd.csv")
dataset_samengevoegd <- raw_inkomen %>%
  left_join(raw_energie, by = "Jaar")
colnames(raw_energie)
colnames(raw_inkomen)
```

## 2.2 Short summary of the data sets

The data sets utilized in this study consist of official statistics sourced from Statistics Netherlands (CBS). The first data set contains average consumer energy tariffs, while the second data set comprises Dutch household income levels across the years 2021, 2022, and 2023, broken down by province.

A limitation of the first data set is that it does not capture the exact monthly or annual energy usage of individual households. However, it provides a reliable overview of the national average fixed and variable tariffs for both natural gas and electricity.

The second data set includes detailed information on standardized, median, and disposable household incomes. Utilizing these specific metrics ensures the most reliable and realistic representation of household financial capacity across the different Dutch provinces.

# Part 3: Quantifying

# Data cleaning

During the import of the Excel tables from the CBS website, it was found that the names of the columns did not match. The name of the first column in the Income datasheet was "jaar", while the name of the first column in the Energie datasheet was "perioden". That issue caused an error in R, it was solved by naming both first columns "jaar". The energy data set was also cleaned by deleting the loose months and use the years instead. After these adjustments the data sets were merged by using left_join(), and R could read the files correctly.

# Describing variabels

# Creation of the new variables

```{r}
# We berekenen de variabelen én hernoemen de oude kolommen in één stap!
dataset_berekend <- dataset_samengevoegd %>%
  mutate(
    # 1. We geven het inkomen meteen een kortere Engelse naam
    income_median = `Inkomen|Mediaan besteedbaar inkomen 1 000 euro`,
    
    # 2. Total gas costs calc (met de originele CBS-namen)
    gas_costs = `Aardgas transporttarief inc BTW Euro/jaar` + 
                `Aardgas leverings tarief vast inc btw Euro/jaar` + 
                (1200 * (`Aardgas|Variabel leveringstarief inc btw Euro/m3` + `aardgas energiebelasting inc btw/m3`)),
    
    # 3. Total electricity costs calc (met de originele CBS-namen)
    electricity_costs = `elektriciteit transporttarief inc btwEuro/jaar` + 
                         `elek vast leveringstarief inc btwEuro/jaar` + 
                         (2500 * (`Elektriciteit|Variabel leveringstarief contractprijs inc btwEuro/kWh` + `Elektriciteit|Energiebelasting inc btwEuro/kWh`)),
    
    # 4. Total energy costs combined
    total_energy_costs = gas_costs + electricity_costs,
    
    # 5. energy_burden (totale energiekosten)
    energy_burden = (total_energy_costs / (income_median * 1000)) * 100,
    
    # 6. gas_burden (pure gaskosten)
    gas_burden = (gas_costs / (income_median * 1000)) * 100,
    
    # 7. electricity_burden (pure stroomkosten)
    electricity_burden = (electricity_costs / (income_median * 1000)) * 100
  )
```

# Explanation new variables

The new variables for this study are:

Energy_burden, represents the portion of the Dutch median disposable income that is spent on total energy costs (gas and electricity combined).

Gas_burden, represents the portion of the Dutch median disposable income that is spent on gas costs. This variable is calculated by adding the variable and fixed costs of gas based on a consumption of 1200m3 per year, the CBS guidelines show that 1200m3 is the normal consumption for a household per year.

Electricity_burden, represents the portion of the Dutch median disposable income that is spent on electricity costs. This variable is calculated by adding the variable and fixed costs of electricity based on a consumption of 2500kWh per year, the CBS guidelines show that 2500kWh is the normal consumption for a household per year(CBS,2023).

# Data Visualisation

# Figrure 1

```{r}
# 1. Data preperation
temporal_data <- dataset_berekend %>% 
  filter(Regios == "Nederland")

ggplot(temporal_data, aes(x = factor(Jaar), y = energy_burden, group = 1)) +
  geom_line(color = "darkblue", linewidth = 1.3) +
  geom_point(color = "red", size = 4) +
  labs(
    title = "1. Temporal Analysis: Development of Energy Burden (2021-2023)",
    subtitle = "National average of energy costs as % of disposable income",
    x = "Year",
    y = "Energy Burden (%)"
  ) +
  theme_minimal()
```

#Figure 2

```{r}
spatial_data <- dataset_berekend %>% 
  filter(Jaar == 2023 & grepl("\\(PV\\)", Regios))

ggplot(spatial_data, aes(x = reorder(Regios, energy_burden), y = energy_burden, fill = energy_burden)) +
  geom_col() +
  coord_flip() + # Horizontaal maken voor leesbaarheid regio's
  scale_fill_gradient(low = "yellow", high = "darkred") +
  labs(
    title = "2. Spatial Analysis: Regional Energy Burden in 2023",
    subtitle = "Comparison between Dutch provinces",
    x = "Province",
    y = "Energy Burden (%)",
    fill = "Burden %"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```

#Figure 3

```{r}
sub_population_data <- dataset_berekend %>% 
  filter(Jaar == 2023) %>% 
  mutate(Type = case_when(
    Regios == "Nederland" ~ "National Average",
    grepl("\\(PV\\)", Regios) ~ "Provinces",
    TRUE ~ "Major Cities" # De rest (zoals Amsterdam) zijn steden
  ))

ggplot(sub_population_data, aes(x = Type, y = energy_burden, fill = Type)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 2, alpha = 0.6) +
  labs(
    title = "3. Sub-population Analysis: Cities vs. Provinces (2023)",
    subtitle = "Comparing the spread of energy burden across different regional types",
    x = "Sub-population Type",
    y = "Energy Burden (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
event_data <- dataset_berekend %>% 
  filter(Regios != "Nederland")

ggplot(event_data, aes(x = factor(Jaar), y = energy_burden)) +
  geom_boxplot(fill = "lightcoral", alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "darkred") +
  # Voeg een visuele lijn/indicator toe die het 'Event' markeert (de start van de energiecrisis begin 2022)
  geom_vline(xintercept = 1.5, linetype = "dashed", color = "black", linewidth = 1) +
  annotate("text", x = 1.6, y = max(event_data$energy_burden) - 2, 
           label = "Start Energy Crisis (2022)", hjust = 0, fontface = "italic", color = "black") +
  labs(
    title = "4. Event Analysis: Impact of the 2022 Energy Crisis",
    subtitle = "Shift in household energy burden before and after the price shock",
    x = "Year",
    y = "Energy Burden (%)"
  ) +
  theme_minimal()
```

# Discussion

Our findings state that, within the limitations mentioned earlier, the share of the Dutch median income spent on energy bills increased significantly each year between 2021 and 2023. It is important to note that precise figures are difficult to determine. However, the fact remains that standard disposable income has not increased as rapidly as energy costs. This trend highlights the growing importance of policies aimed at improving energy affordability and protecting vulnerable households in the Netherlands.
