---
title: "Energy poverty in the Netherlands"
author: "Matthijs, David, Wessel, Quint, Zakaria"
date: "2026-06-08"
output: html_document
---


```{r setup, include=FALSE}
# Hier laden we alle pakketten die nodig zijn voor de opdracht
require(tidyverse)
require(rmarkdown)
require(yaml)
library(tidyr)
library(writexl)
library(readxl)
```
# Part 1: Identify a Social Problem
## 1.1 Describe the Social Problem
In recent years, energy prices in the Netherlands have increased significantly, leading to a substantial rise in fixed costs of Dutch households. This increase in fixed expenses reduces the financial resources available for those household, limiting their capacity to meet other essential needs. To effectively address this growing issue among those households, it is crucial to quantify and understand the extent to which rising energy prices drive poverty rates among the Dutch households. 

# Part 2: Data Sourcing
## 2.1 Load in the data
```{r}
# We laden de dataset rechtstreeks in vanaf een betrouwbare online bron (CBS StatLine)

dataset <- read_csv2("data1.csv")
```
## 2.2 Provide a short summary of the dataset(s)

```{r}
# Hiermee laten we de eerste 6 rijen van de dataset zien
head(dataset)
```

## 2.3 Describe the type of variables included

# Part 3: Quantifying
## 3.1 Data cleaning
## 3.2 Generate necessary variables
## 3.3 Visualize temporal variation
## 3.4 Visualize spatial variation
## 3.5 Visualize sub-population variation
## 3.6 Event analysis

# Part 4: Discussion
## 4.1 Discuss your findings

# Part 5: Reproducibility
## 5.1 Github repository link
## 5.2 Reference list

